---
title: "Elo against the ATP ranking"
subtitle: "Where the two disagree, and who each one rates highest"
author: "Justus Treeger"
date: today
format:
  html:
    toc: true
    df-print: kable
    fig-width: 9
    fig-height: 5
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
---

```{r setup}
suppressMessages(pkgload::load_all("..", quiet = TRUE))
library(ggplot2)

MATCH_DIR   <- "../data-raw/csv/matches"
RANK_DIR    <- "../data-raw/csv/rankings"
ACTIVE_DAYS <- 365
MIN_OVERALL <- 20
MIN_SURFACE <- 10
PAIRING     <- "individual"
LEVELS      <- c("G", "M", "A", "F", "O")

match_files <- list.files(MATCH_DIR, pattern = "^atp_matches_[0-9]{4}\\.csv$",
                          full.names = TRUE)
rank_files  <- list.files(RANK_DIR, pattern = "^atp_rankings_.*\\.csv$",
                          full.names = TRUE)
stopifnot(length(match_files) > 0, length(rank_files) > 0)

raw      <- load_matches(match_files)
matches  <- prepare_matches(raw, levels = LEVELS, retirements = "include")
rankings <- prepare_rankings(load_rankings(rank_files))

fit <- fit_elo(matches, pairing = PAIRING)
AS_OF <- fit$as_of

players <- utils::read.csv(file.path(RANK_DIR, "atp_players.csv"),
                           colClasses = "character")
players$name <- trimws(paste(players$name_first, players$name_last))
nm <- function(id) {
  i <- match(as.character(id), players$player_id)
  ifelse(is.na(i), as.character(id), players$name[i])
}

manifest <- utils::read.csv(file.path(MATCH_DIR, "MANIFEST.csv"))
```

## What this analysis fixes before it starts

Every number below depends on these choices. They are analysis decisions, made
here and visible, not hidden package defaults.

```{r parameters}
data.frame(
  Parameter = c("Match data cut-off", "ATP ranking used", "Tournament levels",
                "Pairing rule", "Activity window", "Minimum matches (overall)",
                "Minimum matches (surface)", "Retirements",
                "K parameters", "Package version",
                "Source", "Source commit", "Files retrieved"),
  Value = c(
    format(AS_OF),
    format(max(rankings$ranking_date[rankings$ranking_date <= AS_OF])),
    paste(LEVELS, collapse = ", "),
    PAIRING,
    paste(ACTIVE_DAYS, "days"),
    MIN_OVERALL, MIN_SURFACE, "included",
    sprintf("k_start=%g, offset=%g, shape=%g",
            fit$config$k_start, fit$config$offset, fit$config$shape),
    as.character(utils::packageVersion("tennelo")),
    as.character(manifest$source[1]),
    substr(as.character(manifest$source_commit[1]), 1, 12),
    nrow(manifest)
  ),
  check.names = FALSE
)
```

Elo is an estimate inside a disclosed model. It is not a measurement of true
playing strength, and nothing below should be read as one.

## 1. Data coverage and exclusions

```{r coverage}
rep <- exclusion_report(matches)
dropped <- rep$dropped[rep$dropped > 0]
data.frame(
  Item = c("Raw rows read", "Rated", names(dropped)),
  Rows = c(rep$n_input, rep$n_kept, unname(dropped)),
  check.names = FALSE
)
```

```{r coverage-check}
stopifnot(rep$n_kept + sum(rep$dropped) == rep$n_input)
cat(sprintf("Rated matches: %s, covering %s to %s, %s players.\n",
            format(fit$n_matches, big.mark = ","),
            fit$first_date, fit$as_of,
            format(fit$n_players, big.mark = ",")))
```

Every excluded row is counted by reason, and the counts reconcile with the
input. No filtering happens silently.

## 2. Elo leaderboard, overall

```{r leaderboard}
lb <- elo_leaderboard(fit, n = 20, active_within = ACTIVE_DAYS,
                      min_matches = MIN_OVERALL)
lb$Player <- nm(lb$player_id)
lb$Rating <- round(lb$rating, 1)
lb[, c("rank", "Player", "Rating", "matches", "last_match")]
```

## 3. By surface

```{r surface-fits}
SURFACES <- c("Hard", "Clay", "Grass")
surface_fits <- lapply(SURFACES, function(s) fit_elo(matches, surface = s,
                                                     pairing = PAIRING))
names(surface_fits) <- SURFACES

surface_tops <- lapply(SURFACES, function(s) {
  f <- surface_fits[[s]]
  t <- elo_leaderboard(f, n = 10, as_of = AS_OF,
                       active_within = ACTIVE_DAYS, min_matches = MIN_SURFACE)
  data.frame(Surface = s, Rank = t$rank, Player = nm(t$player_id),
             Rating = round(t$rating, 1), Matches = t$matches)
})
do.call(rbind, surface_tops)
```

```{r surface-sizes}
data.frame(
  Surface = SURFACES,
  Matches = vapply(surface_fits, function(f) f$n_matches, numeric(1)),
  Players = vapply(surface_fits, function(f) f$n_players, numeric(1))
)
```

Grass carries far fewer matches than hard or clay, so its ratings are the least
stable of the three and should be read with that in mind.

Carpet is left out. It was a real surface for decades -- 22% of matches in 1975
-- but the ATP dropped it at the end of the 2000s, and no currently active
player has a meaningful carpet record. The matches stay in the overall model;
a carpet leaderboard of active players would be empty of meaning.

## 4. Elo against the ATP ranking

```{r comparison}
cmp <- compare_rankings(fit, rankings, as_of = AS_OF,
                        active_within = ACTIVE_DAYS,
                        min_matches = MIN_OVERALL, n = 20)
data.frame(
  Measure = c("Elo cut-off", "ATP ranking date", "Same day?",
              "Players compared", "Matched on player id",
              "Spearman rank correlation", "Top-20 overlap"),
  Value = c(format(cmp$elo_date), format(cmp$atp_date),
            if (cmp$same_day) "yes" else sprintf("no, %.0f days apart", cmp$gap_days),
            cmp$n_elo, sprintf("%d (%.1f%%)", cmp$n_matched, 100 * cmp$pct_matched),
            sprintf("%.3f", cmp$spearman),
            sprintf("%d of %d", cmp$top_n_overlap, cmp$top_n)),
  check.names = FALSE
)
```

```{r scatter}
cc <- cmp$comparison
cc$Player <- nm(cc$player_id)
ggplot(cc, aes(atp_rank, elo_rank)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_point(alpha = 0.55, size = 1.6) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "ATP rank (log)", y = "Elo rank (log)",
       title = "Where the two rankings agree, and where they do not",
       subtitle = sprintf("%d active players, %s. Points above the line are rated higher by ATP.",
                          cmp$n_matched, format(AS_OF))) +
  theme_minimal()
```

## 5. The largest disagreements

```{r divergence-up}
up <- utils::head(cc[order(-cc$rank_diff), ], 10)
data.frame(Player = up$Player, `Elo rank` = up$elo_rank,
           `ATP rank` = up$atp_rank, Difference = up$rank_diff,
           Rating = round(up$rating, 1), Matches = up$matches,
           `Last match` = up$last_match, check.names = FALSE)
```

```{r divergence-down}
dn <- utils::head(cc[order(cc$rank_diff), ], 10)
data.frame(Player = dn$Player, `Elo rank` = dn$elo_rank,
           `ATP rank` = dn$atp_rank, Difference = dn$rank_diff,
           Rating = round(dn$rating, 1), Matches = dn$matches,
           `Last match` = dn$last_match, check.names = FALSE)
```

The two tables have distinct characters, and both trace back to a documented
limitation of this model rather than to noise.

Players Elo rates far above their ATP rank are largely returning from long
absences. An ATP ranking decays when a player stops entering tournaments; an
Elo rating does not move at all. The model has no absence correction, so a
player who was strong, disappeared, and came back keeps the rating they left
with.

Players ATP rates far above their Elo rank are largely young players climbing
fast. They built much of their ranking at Challenger level, which this analysis
excludes, so they enter the data at 1500 with no earned history. Their Elo has
not yet caught up with what they have actually done.

## 6. Careers

```{r careers}
top6 <- utils::head(lb$player_id, 6)
plot_elo_history(fit, top6) +
  labs(title = "Rating histories of the current top six")
```

```{r peaks}
peaks <- do.call(rbind, lapply(top6, function(p) elo_peak(fit, p, min_matches = MIN_OVERALL)))
data.frame(Player = nm(peaks$player_id), `Peak rating` = round(peaks$rating, 1),
           `Reached at` = peaks$tourney_date, Matches = peaks$matches_played,
           check.names = FALSE)
```

Peak dates are tournament start dates. The source records when a tournament
began, not when a match was played, so that is the finest resolution available.

## 7. Sensitivity

Every headline above rests on choices. These are the same numbers under
plausible alternatives.

```{r sensitivity-pairing}
alt <- fit_elo(matches, pairing = "symmetric")
alt_lb <- elo_leaderboard(alt, n = 20, active_within = ACTIVE_DAYS,
                          min_matches = MIN_OVERALL)
both <- merge(
  data.frame(player_id = lb$player_id, individual = lb$rank),
  data.frame(player_id = alt_lb$player_id, symmetric = alt_lb$rank),
  by = "player_id", all = TRUE
)
both$Player <- nm(both$player_id)
both <- both[order(both$individual), c("Player", "individual", "symmetric")]
both
```

```{r sensitivity-table}
variants <- list(
  `baseline`                = list(pairing = PAIRING, ret = "include", act = ACTIVE_DAYS, min = MIN_OVERALL),
  `symmetric pairing`       = list(pairing = "symmetric", ret = "include", act = ACTIVE_DAYS, min = MIN_OVERALL),
  `retirements excluded`    = list(pairing = PAIRING, ret = "exclude", act = ACTIVE_DAYS, min = MIN_OVERALL),
  `activity window 180d`    = list(pairing = PAIRING, ret = "include", act = 180, min = MIN_OVERALL),
  `minimum 50 matches`      = list(pairing = PAIRING, ret = "include", act = ACTIVE_DAYS, min = 50)
)
rows <- lapply(names(variants), function(v) {
  cfg <- variants[[v]]
  mm <- if (cfg$ret == "include") matches else
    prepare_matches(raw, levels = LEVELS, retirements = "exclude")
  ff <- if (cfg$pairing == PAIRING && cfg$ret == "include") fit else
    fit_elo(mm, pairing = cfg$pairing)
  cm <- compare_rankings(ff, rankings, as_of = AS_OF,
                         active_within = cfg$act, min_matches = cfg$min, n = 20)
  tb <- elo_leaderboard(ff, n = 1, active_within = cfg$act, min_matches = cfg$min)
  data.frame(Variant = v, `Top player` = nm(tb$player_id),
             Spearman = round(cm$spearman, 3),
             `Top-20 overlap` = cm$top_n_overlap,
             Compared = cm$n_elo, check.names = FALSE)
})
do.call(rbind, rows)
```

```{r sensitivity-k}
k_rows <- lapply(list(c(150, 5, 0.4), c(250, 5, 0.4), c(400, 5, 0.4), c(250, 5, 0.2)),
  function(k) {
    ff <- fit_elo(matches, pairing = PAIRING,
                  k_start = k[1], offset = k[2], shape = k[3])
    cm <- compare_rankings(ff, rankings, as_of = AS_OF,
                           active_within = ACTIVE_DAYS, min_matches = MIN_OVERALL)
    tb <- elo_leaderboard(ff, n = 1, active_within = ACTIVE_DAYS,
                          min_matches = MIN_OVERALL)
    data.frame(`k_start` = k[1], offset = k[2], shape = k[3],
               `Top player` = nm(tb$player_id),
               Spearman = round(cm$spearman, 3),
               `Top-20 overlap` = cm$top_n_overlap, check.names = FALSE)
  })
do.call(rbind, k_rows)
```

## 8. What this cannot tell you

- **The order within a week is approximate.** `tourney_date` is the tournament
  start date, not the match date, so matches played in the same week can be
  rated out of sequence. Player ids break remaining ties deterministically;
  that makes the result reproducible, not chronologically true.
- **Challenger and qualifying results are excluded.** Everyone enters at 1500,
  which is roughly an average tour professional. A real newcomer is below that
  and hands out points they never earned. This is visible in section 5.
- **Early ratings are uncertain.** A player near the minimum match threshold
  can sit high on a short run of upsets. `min_matches` limits this; it does not
  remove it.
- **Grass has a small sample**, and carpet is excluded from the leaderboards
  entirely.
- **There is no absence or injury correction.** A player returning after a year
  out keeps the rating they left with. This is the main driver of section 5's
  first table.
- **Surface and overall ratings are not blended.** Each surface model is fitted
  independently and knows nothing of the others.
- **Cross-era comparison is out of scope.** Under `individual` pairing the pool
  drifts, so ratings from different decades are not on a common scale. Every
  comparison here is between players on one date.
- **Elo is a model estimate, not a measurement.** A different K, a different
  match set or a different pairing rule gives different numbers, as section 7
  shows directly.
